Explanation of How to Do An ANOVA
===================================

```{r data-creation}
balloon.data <- data.frame(color=c(rep("green",6), rep("red",12), rep("blue",6)),
                          shape = rep(c("square", "circle"),12),
                          size = rpois(24, lambda=5)*abs(rnorm(24)))
```

Check Distribution of Your Data
---------------------------------

First check the distribution of your groups of data using a Shapiro-Wilk Test and then then the presumption of equal variance with a Levene's Test

```{r shapiro-test}
#super duper easy shapiro test shortcut
with(balloon.data, tapply(size, shape, FUN=function(x)shapiro.test(x)$p.value))
with(balloon.data, tapply(size, list(color,shape), FUN=function(x)shapiro.test(x)$p.value))
```

One Way ANOVA
--------------
This is used when you have one variable, with several types.  In this example you have several colors of balloons and are testing the hypothesis that size is affected by balloon color.

```{r one-way-anova}
#one way anova (two types)
summary(aov(size~shape, data=balloon.data))
#same as a t-test
t.test(size~shape, data=balloon.data, var.equal=T)

#one way anova (3 types)
summary(aov(size~color, data=balloon.data))
```

Two-Way ANOVA
---------------
This is used when you have two (or more) potential variables which affect your dependent variable.

```{r two-way-anova}
#two way anova, with interaction
summary(aov(size~color*shape, data=balloon.data))

#two way anova, without interaction
summary(aov(size~color+shape, data=balloon.data))
```

Assumptions of ANOVA Analyses
-------------------------------
Although when group sizes are equal, ANOVA is quite robust to non-normality and non-equal variance, you should test these assumptions.  Normality was tested above.  If the data is not normal, you could transform your dependent variable (ie look at log(size) rather than size).  You could also use a non-parametric test such as a Kruskal-Wallis test

```{r non-parametric-tests}
#back to the levene's test
library(car)
with(balloon.data, leveneTest(size~shape*shape))

#if not normally distributed try a kruskal wallis test for non-parametric data
kruskal.test(size~color, data=balloon.data)
#anova for comparason
summary(aov(size~color, data=balloon.data))
```

Session Information
--------------------

```{r session-info, echo=FALSE}
sessionInfo()
```