Explanation of How to Do An ANOVA
===================================

```{r data-creation}
balloon.data <- data.frame(color=c(rep("green",6), rep("red",12), rep("blue",6)),
                          shape = rep(c("square", "circle"),12),
                          size = rpois(24, lambda=5)*abs(rnorm(24)))
```

# Check Distribution of Your Data

First check the distribution of your groups of data using a Shapiro-Wilk Test and then then the presumption of equal variance with a Levene's Test

```{r shapiro-test}
#super duper easy shapiro test shortcut
with(balloon.data, tapply(size, shape, FUN=function(x)shapiro.test(x)$p.value))
with(balloon.data, tapply(size, list(color,shape), FUN=function(x)shapiro.test(x)$p.value))
```

# One Way ANOVA

This is used when you have one variable, with several types.  In this example you have several colors of balloons and are testing the hypothesis that size is affected by balloon color.

```{r one-way-anova}
#one way anova (two types)
summary(aov(size~shape, data=balloon.data))
#same as a t-test
t.test(size~shape, data=balloon.data, var.equal=T)

#one way anova (3 types)
summary(aov(size~color, data=balloon.data))
```

# Two-Way ANOVA

This is used when you have two (or more) potential variables which affect your dependent variable.

```{r two-way-anova}
#two way anova, with interaction
summary(aov(size~color*shape, data=balloon.data))

#two way anova, without interaction
summary(aov(size~color+shape, data=balloon.data))
```

## Interactions

For interpreting interactions, if the interaction term is significant then the two main effects are not useful  For a more detailed discussion of interpreting interactions see:

Nieuwenhuis, S., Forstmann, B. U. & Wagenmakers, E.-J. J. Erroneous analyses of interactions in neuroscience: A problem of significance. Nat. Neurosci. 14, 1105–1107 (2011). http://dx.doi.org/10.1038/nn.2886

# Assumptions of ANOVA Analyses

Although when group sizes are equal, ANOVA is quite robust to non-normality and non-equal variance, you should test these assumptions.  Normality was tested above.  If the data is not normal, you could transform your dependent variable (ie look at log(size) rather than size).  You could also use a non-parametric test such as a Kruskal-Wallis test

```{r non-parametric-tests}
#back to the levene's test
library(car)
with(balloon.data, leveneTest(size~shape*shape))

#if not normally distributed try a kruskal wallis test for non-parametric data
kruskal.test(size~color, data=balloon.data)
#anova for comparason
summary(aov(size~color, data=balloon.data))

#test for normality by runnning a shapiro test on the residuals
shapiro.test(residuals(aov(size~color+shape, data=balloon.data)))
```

Post-Hoc Tests
----------------

If you get a significant result from the omnibus ANOVA, you often will want to do further post-hoc testing to look at pairwise comparisons.  There are three main ways to do this:

* Pairwise T-Tests (remember to check the assumptions of normality and equal variance when choosing a test)
* Tukey's Honest Significant Difference -- Compares all combinations
* Dunnett's Test -- Compares all values to a control

```{r post-hoc-tests}
#save an anova object
balloon.aov <- aov(size~color,  data=balloon.data)
#can run a Tukey's test on this
TukeyHSD(balloon.aov)
#this is the same as:
library(multcomp)
summary(glht(balloon.aov, linfct=mcp(color="Tukey")))

#Can also use glht for Dunnett's tests
#define the control group to what you want to compare to
balloon.data$color <- relevel(balloon.data$color, ref="blue")
#run a dunnett's test
summary(glht(balloon.aov, linfct=mcp(color="Dunnett")))
#default is single-step but can manually set FDR adjustments
summary(glht(balloon.aov, linfct=mcp(color="Dunnett")), adjusted(type="BH"))
summary(glht(balloon.aov, linfct=mcp(color="Dunnett")), adjusted(type="bonferroni"))

#to do pairwise t tests can do
pairwise.t.test(balloon.data$size, balloon.data$color)
#default is Holm but can set to whatever you want or none
pairwise.t.test(balloon.data$size, balloon.data$color, p.adjust.method="BH")
pairwise.t.test(balloon.data$size, balloon.data$color, p.adjust.method="none")
```

# Session Information


```{r session-info, echo=FALSE}
sessionInfo()
```